Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-12 examples.tex
more figure updates, scaffolding prose smoothing in 5.1

\todo[inline]{TODO: resolve entry+KERNBASE question from chat, add information about rax holding \lstinline|entry_val & 0x1| and the conditional
outcome that that is 0 in the then branch}

@@

;; setting up the stack

@@

jne    161 <pte_get_next_table_succ+0xa1> (* \label{line:check_entry_present_jump} *) ; Jump if the present bit is not zero

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$
sub    0x10, rdi
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$

@@

Thus the fractional ownership of the physical location is enough for Line \ref{line:read_entry_contents} to access the entry, though in \lstinline|get_next_table|
the caller has pulled that piece of information out of the invariant and passed it for the entry at hand, which is why this appears explicitly in assertions,
and the argument to the invariant is $\Xi\setminus\{\mathsf{entry}\}$\todo{update pending entry+KB} (indexing by the set $\Xi$ allows us to borrow the physical resources
for a specific page table entry out of the invariant, and later put them back).
Line \ref{line:check_entry_present_jump}'s conditional then determines in the fall-through case that the bit is not set, which resolves $\textsf{qfrac} = 1$,
and permits the new entry to be installed (in ellided code around Line \ref{line:install_new_entry})

@@

Concretely speaking, going back to Line 15 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry} while preserving the soundness of memory mappings, our extended invariant introduces the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated in Line 21 in Figure \ref{fig:calltopteinitialize}.
\begin{comment}
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}
\end{comment}
 \subsubsection{Ensuring Presence of an Entry}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 % \[ \begin{array}{l}
 %   \mathsf{and }\; 0x1, \;rdi \\
 %   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
 %   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 % \end{array}
 % \] 
Lines \ref{line:mask_present}--\ref{line:check_entry_present} in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.

>>>>

\todo[inline]{TODO: resolve entry+KERNBASE question from chat, update all rdi points-tos after line \ref{line:pass_addrof_next}}

@@

... ;; setting up the stack

@@

jne    161 <pte_get_next_table_succ+0xa1> (* \label{line:check_entry_present_jump} *) ;; Jump if the present bit is not zero
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val \& 0x1} }_{\rtv}$
$\specline{  \textsf{entry} \mapsto_{\textsf{id}} \textsf{v} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  \ast \texttt{rax}  \mapsto_{\textsf{r}} \textsf{entry\_val \& 0x1} \ast \ulcorner\textsf{entry\_val \& 0x1}=\textsf{0x0}\urcorner}_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow  \lnot(\textsf{entry\_val \& 0x1}=\textsf{0x1})  \urcorner }_{\rtv}$ (*\label{line:before_concluding_qfrac1}*)
sub    0x10, rdi ;; Store the value of rbp minus 16 bytes (address of (*\textsf{next}*)) into rdi (*\label{line:pass_addrof_next}*)

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$ (*\label{line:after_concluding_qfrac1}*)

@@

Thus the fractional ownership of the physical location is enough for Line \ref{line:read_entry_contents} to access the entry, though in \lstinline|get_next_table|
the caller has pulled that piece of information out of the invariant and passed it for the entry at hand, which is why this appears explicitly in assertions,
and the argument to the invariant is $\Xi\setminus\{\mathsf{entry}\}$\todo{update pending entry+KB} (indexing by the set $\Xi$ allows us to borrow the physical resources
for a specific page table entry out of the invariant, and later put them back).
Line \ref{line:check_entry_present_jump}'s conditional then determines in the fall-through case that the bit is not set, which 
together with other facts entails $\textsf{qfrac} = 1$ at Line \ref{line:after_concluding_qfrac1},
and permits storing a new entry (in ellided code around Line \ref{line:install_new_entry})

@@

Concretely speaking, going back to Line 15 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry} while preserving the soundness of memory mappings, our extended invariant introduces the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated in Line 21 in Figure \ref{fig:calltopteinitialize}.
\begin{comment}
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}
\end{comment}

 \subsubsection{Ensuring Presence of an Entry}
\todo[inline]{some of this content is now earlier, rename to ``installing a new entry'' and just talk about the invariant update from installing the newly-allocated
entry. include explanation of the resources returned from initialization call.
Carefully check speck and then explain code of pte initialize, which currently has not prose (though this is a candidate for appendix in an oopsla submission)
}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 % \[ \begin{array}{l}
 %   \mathsf{and }\; 0x1, \;rdi \\
 %   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
 %   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 % \end{array}
 % \] 
Lines \ref{line:mask_present}--\ref{line:check_entry_present} in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.
